---
title: "Analysis"
format: html
---

## Goal of this notebook

Analyzing Austin weather data, specifically temperature and precipitation.  

## Setup

Loading the libraries.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
```

## Import the data

Importing the weather data into the project.

```{r}
#| label: import

weather <- read_rds("data-processed/01-weather-cleaned.rds")

weather |> glimpse()
```

## Days with most rain

Which days had the most rain, and how much rain are we talking about?

```{r}
weather |>
  select(date, prcp) |>
  arrange(desc(prcp)) |>
  filter(prcp >= 5)
```

Data Takeaway: The heaviest rainfall days in Austin brought more than five inches of rain in a single day, with the highest reaching 7.55 inches on Nov. 15, 2001.

## Hottest days

What are the hottest days in history?

```{r}
weather |>
  select(date, tmax) |>
  arrange(desc(tmax)) |>
  head(10)
```

Data Takeaway: Austin’s hottest recorded temperature was 112 degrees, reached on Sept. 5, 2000, and Aug. 28, 2011. Multiple hot days occurred in 2011, appearing three times among the top 10 hottest days.

## Coldest days

What are the coldest days in history? 

```{r}
weather |>
  select(date, tmin) |>
  arrange(tmin) |>
  head(10)
```

Data Takeaway: The coldest days in Austin’s recorded history reached as low as -2 degrees on Jan. 31, 1949.

## Days over 100

Which years had the most days of 100+ temperature? i.e., group by year and count the number of days with 100+ temps.

```{r}
weather |>
  filter(tmax >= 100) |>
  group_by(year = year(date)) |>
  summarize(days_100_plus = n()) |>
  arrange(desc(days_100_plus)) |>
  filter(days_100_plus >= 50)
```

Data Takeaway: In 2011, Austin experienced 90 days with temperatures over 100 degrees, beating out 2023 by 10 more days over 100 degrees.

