---
title: "Analysis"
format: html
---

## Goal of this notebook

Analyzing Austin weather data, specifically temperature and precipitation.  

## Setup

Loading the libraries.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(lubridate)
```

## Import the data

Importing the weather data into the project.

```{r}
#| label: import

weather <- read_rds("data-processed/01-weather-cleaned.rds")

weather |> glimpse()
```

## Days with most rain

Which days had the most rain, and how much rain are we talking about?

```{r}
weather |>
  select(date, prcp) |>
  arrange(desc(prcp)) |>
  filter(prcp >= 5)
```

Data Takeaway: The heaviest rainfall days in Austin brought more than five inches of rain in a single day, with the highest reaching 7.55 inches on Nov. 15, 2001.

## Hottest days

What are the hottest days in history?

```{r}
weather |>
  select(date, tmax) |>
  arrange(desc(tmax)) |>
  head(10)
```

Data Takeaway: Austin’s hottest recorded temperature was 112 degrees, reached on Sept. 5, 2000, and Aug. 28, 2011. Multiple hot days occurred in 2011, appearing three times among the top 10 hottest days.

## Coldest days

What are the coldest days in history? 

```{r}
weather |>
  select(date, tmin) |>
  arrange(tmin) |>
  head(10)
```

Data Takeaway: The coldest days in Austin’s recorded history reached as low as -2 degrees on Jan. 31, 1949.

## Days over 100

Which years had the most days of 100+ temperature? i.e., group by year and count the number of days with 100+ temps.

```{r}
weather |>
  filter(tmax >= 100) |>
  group_by(year = year(date)) |>
  summarize(days_100_plus = n()) |>
  arrange(desc(days_100_plus)) |>
  filter(days_100_plus >= 50)
```

Data Takeaway: In 2011, Austin experienced 90 days with temperatures over 100 degrees, beating out 2023 by 10 more days over 100 degrees.

## Years with the most rainfall

Which years had the most total rainfall and how much?

```{r}
weather |>
  group_by(year = year(date)) |>
  summarize(total_rain = sum(prcp, na.rm = TRUE)) |>
  arrange(desc(total_rain))
```

Data Takeaway: Austin’s wettest year in the dataset was 2015, which recorded nearly 60 inches of rainfall. Other notably rainy years include 2004, 1991 and 1957, all exceeding 50 inches.

## Years with the least rainfall

Which years had the lowest total rainfall?

```{r}
weather |>
  group_by(year = year(date)) |>
  summarize(total_rain = sum(prcp, na.rm = TRUE)) |>
  arrange(total_rain) |>
  head(10)
```

Data Takeaway: The dataset shows 2026 with only 1.44 inches of rainfall so far, but this year is still ongoing, so the total isn’t final. Excluding 2026, the driest complete years include 1938, 1954 and 1956, all under 16 inches.

## Years with the most snowfall

Which years had the most total snowfall?

```{r}
weather |>
  group_by(year = year(date)) |>
  summarize(total_snow = sum(snow, na.rm = TRUE)) |>
  arrange(desc(total_snow))
```

Data Takeaway: The snowiest year was 1985 with 8.7 inches, followed by 2021 with 7.9 inches and 1944 with 7.0 inches. Most other years recorded only a few inches, showing that significant snow is uncommon in Austin.

## Average high and low temperature by year

Calculating the average high temperature and average low temperature for each year.

```{r}
weather |>
  group_by(year = year(date)) |>
  summarize(
    avg_high_temp = mean(tmax, na.rm = TRUE),
    avg_low_temp = mean(tmin, na.rm = TRUE)
  ) |>
  arrange(desc(avg_high_temp))
```

Data Takeaway: The warmest years in Austin, based on average highs and lows, include 1938 and 2011, both with average highs above 84 degrees. Average lows for these years also remain relatively high, although 2024 has the highest low temperature, just over 62 degrees. 

## Average monthly rainfall since 1990

Calculating total rainfall for each month since 1990, then average rainfall for each month.

```{r}
rain_month_year <- weather |>
  filter(year(date) >= 1990) |>
  group_by(year = year(date), month = month(date, label = TRUE)) |>
  summarize(total_rain = sum(prcp, na.rm = TRUE))

rain_month_year |>
  ungroup() |>
  group_by(month) |>
  summarize(avg_monthly_rain = mean(total_rain, na.rm = TRUE)) |>
  arrange(desc(avg_monthly_rain))
```

Data Takeaway: Since 1990, May has been the wettest month in Austin by a large margin, averaging just over five inches of rain. February is the least rainy month, being the only one averaging under two inches.

## Average yearly temperature

Calculating the average yearly temperature using the average high and average low temperatures.

```{r}
weather |>
  group_by(year = year(date)) |>
  summarize(
    avg_high_temp = mean(tmax, na.rm = TRUE),
    avg_low_temp = mean(tmin, na.rm = TRUE)
  ) |>
  mutate(avg_year_temp = (avg_high_temp + avg_low_temp) / 2) |>
  arrange(desc(avg_year_temp))
```

Data Takeaway: Austin’s warmest year based on average yearly temperature is 1938, but all the top five years exceed 72 degrees. Disregarding 2026, 1940 was notably the lowest average yearly temperature, just under 66 degrees. 

## Earliest 100-degree day each year

What is the earliest date each year that reached 100 degrees?

```{r}
weather |>
  filter(tmax >= 100) |>
  group_by(year = year(date)) |>
  arrange(date) |>
  slice_min(date, n = 1) |>
  select(year, date, tmax) |>
  ungroup() |>
  arrange(date)
```

Data Takeaway: The earliest 100-degree days in Austin vary widely by year. For example, in 1938, the first 100-degree day was on Aug. 15, while in 1939 it occurred as early as June 15.

Which year had the earliest date?

```{r}
weather |>
  filter(tmax >= 100) |>
  mutate(
    day_of_year = yday(date),
    year = year(date)
  ) |>
  slice_min(day_of_year, n = 1) |>
  select(year, date, tmax)
```

Data Takeaway: The earliest 100-degree day in Austin’s recorded history occurred on May 4, 1984, reaching 100 degrees.  

## Latest freeze before July 1

Which year had the latest freeze date?

```{r}
weather |>
  filter(tmin <= 32, month(date) < 7, year(date) < 2026) |> 
  mutate(year = year(date)) |>
  slice_max(date, n = 1) |>
  select(year, date, tmin)
```

Data Takeaway: The latest freeze in Austin before July occurred on Feb. 22, 2025, with a low of 30 degrees.

